\myTitle

### 2015-02-10 tex/main-body-text.tex

\section*{Comparison of {\sc structure} and {\sc flock}}
{\sc flock}, initially designed to map admixture in hybrid populations 
\citep{Duc&Tur2009}, was recently proposed as an alternative to 
{\sc structure} to estimate the number of populations from a collection with 
unsampled source populations \citep{Duc&Tur20123}. In doing so, 
{\sc flock} relies on individuals with like genotypes clustering together 
through an iterative process. It stands to reason then, a reasonable way 
to gauge how well the algorithm accomplishes the main objective 
clustering of individuals into correct groups is to simulate data of known 
population structure with varying levels of differentiation and 
evaluate how well it does assuming the correct number of populations. 
The implementation of {\sc flock} as a VBA procedure (macro) in 
Microsoft Excel precludes a rigorous evaluation of its behavior due 
to the inability to efficiently batch process. Consequenty, we adapted 
the source code from {\sc structure} as describe above to create {\sc flockture}
(code provided on GitHub repository https://github.com/eriqande/flockture) 
in order to accurately assess the strengths and weaknesses of the {\sc flock} 
algorithm. First, using simulated data we demonstrated that {\sc flock} 
and {\sc flockture} are essentially equivalent programs.
Second, we simulated and analyzed datasets with {\sc flockture} 
and three models in {\sc structure} (the admixture and correlated 
allele frequency (A-C) model, the no-admixture and correlated 
allele frequency (NA-C) model, and the no-admixture and non-correlated 
allele frequency (NA-NC) model) to help better understand
under what conditions, if any, running {\sc flock} may be preferable to {\sc structure} .

@@

To compare {\sc flock} and {\sc flockture} 4 datasets composed of 5 populations
were simulated using a coalescent based program (code provided on GitHub repository 
https://github.com/eriqande/flock-comment/simdata). Each of the 5 populations
were comprised of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes
and an average pairwise $F_{ST}$ of 0.03, 0.02, 0.01, and 0.005 among
populations. Both programs were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into corrected population as defined by a zero-one loss function, 
(2) individual $q_i$ values, and (3) plateau records. All runs were performed on a JNCS D685 with an i7 3.40GHz 
processor and 8GB of RAM running 64-bit Windows 7.

Once we were confident that no major differences in inference existed between  {\sc flock} and {\sc flockture}
we used  {\sc flockture} to analyze 48 datasets for two marker types 
(microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader 
range of population differentiation. The 96 datasets were simulated as above (five populations 
comprised of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes for the micro satellites
and 96 loci for the SNPs)
with average pairwise $F_{ST}$ values ranging between 0.005 and 0.040. 
 {\sc flockture} was run with the default run parameters and each run of 
{\sc structure} consisted of a 5,000 sweeep burn-in followed by a 20,000 sweep sample
from the posterior distribution. All runs were performed on a Macbook with a 2.3 GHz i7 processor
and 16GB of RAM running OS X ver. 10.7.5. The program {\sc clumpp} \citep{Jak&Ros2007} 
was used to relabel clusters among different runs,
while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.

\subsection*{{\sc flock} and {\sc flockture}} 
{\sc flock} and {\sc flockture} provided similar clustering of individuals among all four 
simulated datasets. The zero-one loss across all datasets had eqivalent results Fig.~\ref{fig:FvFloss}.
When genetic differentiation is high both programs correctly apportion the majority of individuals and 
as genetic differentiation degrades among populations both programs show corresponding declines in 
assignment ability. Neither the mean nor range of loss between {\sc flock} and {\sc flockture} differs
substantially among the simulated datasets. These assignments were dependent on the normalized likelihood
values ($q_i$), which while they were comparable may be slightly biased toward the bounds (0,1)
for {\sc flock}. In an attempt to understand the deviations being observed we took a subset of the 
first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
{\sc flock} with all individuals coming from population 1 starting in reference population 1 and all 
individuals from population 2 starting in reference population 2. {\sc flock} was run a second time
with the same starting clusters, but with the reference population labels switched. While the final allocation
of individuals matched the mean LLOD was slightly different (5.045009 vs. 5.028084). As a result, comparisons
between {\sc flock} and {\sc flockture} were limited to random initial starting points and should not 
be identical, but should have displayed equivalent behavior. Plateaus were only observed for the dataset 
with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although 
the length of each plateau record was slightly longer for {\sc flockture} and on average had more runs converging on
the largest plateau (Supp. Table 1). Multiple runs of {\sc flockture} demonstrated that the differences are well within the
variability that we would expect from randomizing the initial grouping of individuals to 
initiate the algorithms.

@@

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well differentiated populations. Both programs have less than 1\% assignment error for most runs 
at values of $F_{ST}  \geq$ 0.0603 (Fig.~\ref{fig:FvSloss}). Infrequently, the 
NA-NC and NA-C models in {\sc structure} grouped two populations as one and 
assign individuals from another to two separate populations (Fig.~\ref{fig:FvSdistruct}  Simulation 1). This phenomena 
disappears as population differentiation declines. As the subtlety of genetic structure 
increased loss for both programs increases. 
Performance of the two programs as $F_{ST}$ approaches 0.0238 is relatively similar
characterized by the A-C model in {\sc structure} having the largest loss, the NA-C and NA-NC models performing
relatively similar and {\sc flockture} fluctuating around the results of the other three models. The scale of these 
differences is extremely small - on the order of 10 individuals being misassigned. As genetic 
differentiation continues to decline {\sc flockture}  showed consistently better performance 
at levels of $F_{ST}$ between 0.0238 to 0.0229.  Again, these differences are on the scale of roughly 20 individuals
or 1\%. In simulation 12 we observed the variation in {\sc flockture} results increase drastically (Fig.~\ref{fig:FvSloss}),
and for all other simulated datasets {\sc structure} had a lower assignment error.

>>>>

\section*{Comparison of {\sc structure} and {\sc flock}}
{\sc flock}, initially designed to map admixture in hybrid populations 
\citep{Duc&Tur2009}, was recently proposed as an alternative to 
{\sc structure} for estimating the number of populations from a collection with 
unsampled source populations \citep{Duc&Tur2012}. In doing so, 
{\sc flock} relies on individuals with like genotypes clustering together 
through an iterative process. It stands to reason then, a reasonable way 
to gauge how well the algorithm accomplishes the main objective 
of clustering individuals into their correct groups is to simulate data of known 
population structure with varying levels of differentiation and 
evaluate how well it does assuming the correct number of populations. 
The implementation of {\sc flock} as a VBA procedure (macro) in 
Microsoft Excel precludes a rigorous evaluation of its behavior due 
to the inability to efficiently batch process. Consequenty, we adapted 
the source code from {\sc structure} as describe above to create {\sc flockture}
(code provided on GitHub repository https://github.com/eriqande/flockture) 
in order to accurately assess the strengths and weaknesses of the {\sc flock} 
algorithm. First, using simulated data we demonstrated that {\sc flock} 
and {\sc flockture} are essentially equivalent programs.
Second, we simulated and analyzed datasets with {\sc flockture} 
and three models in {\sc structure} (the admixture and correlated 
allele frequency (A-C) model, the no-admixture and correlated 
allele frequency (NA-C) model, and the no-admixture and non-correlated 
allele frequency (NA-NC) model) to help better understand
under what conditions, if any, running {\sc flock} may be preferable to {\sc structure} .

@@

To compare {\sc flock} and {\sc flockture} 4 datasets composed of 5 populations
were simulated using a coalescent based program (code provided on GitHub repository 
https://github.com/eriqande/flock-comment/simdata). Each of the 5 populations
were composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes
and an average pairwise $F_{ST}$ of 0.03, 0.02, 0.01, and 0.005 among
populations. Both programs were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into corrected population as defined by a zero-one loss function, 
(2) individual $q_i$ values, and (3) plateau records. All runs were performed on a JNCS D685 with an i7 3.40GHz 
processor and 8GB of RAM running 64-bit Windows 7.

Once we were confident that no major differences in inference existed between  {\sc flock} and {\sc flockture}
we used  {\sc flockture} to analyze 48 datasets for two marker types 
(microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader 
range of population differentiation. The 96 datasets were simulated as above (five populations 
composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes for the microsatellites
and 96 loci for the SNPs)
with average pairwise $F_{ST}$ values ranging between 0.004 and 0.045. 
 {\sc flockture} was run with the default run parameters and each run of 
{\sc structure} consisted of a 5,000 sweeep burn-in followed by a 20,000 sweep sample
from the posterior distribution. All runs were performed on a Macbook with a 2.3 GHz i7 processor
and 16GB of RAM running OS X ver. 10.7.5. The program {\sc clumpp} \citep{Jak&Ros2007} 
was used to relabel clusters among different runs,
while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.

\subsection*{{\sc flock} and {\sc flockture}} 
{\sc flock} and {\sc flockture} provided similar clustering of individuals among all four 
simulated datasets. The zero-one loss across all datasets had eqivalent results (Fig.~\ref{fig:FvFloss}).
When genetic differentiation is high both programs correctly apportion the majority of individuals and 
as genetic differentiation degrades among populations both programs show corresponding declines in 
assignment ability. Neither the mean nor range of loss between {\sc flock} and {\sc flockture} differs
substantially among the simulated datasets. These assignments were dependent on the normalized likelihood
values ($q_i$), which while they were comparable may be slightly biased toward the bounds (0,1)
for {\sc flock} (Fig.~\ref{fig:FvFqi}). In an attempt to understand the deviations being observed we took a subset of the 
first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
{\sc flock} with all individuals coming from population 1 starting in reference population 1 and all 
individuals from population 2 starting in reference population 2. {\sc flock} was run a second time
with the same starting clusters, but with the reference population labels switched. While the final allocation
of individuals matched the mean LLOD was slightly different (5.045009 vs. 5.028084). As a result, comparisons
between {\sc flock} and {\sc flockture} were limited to random initial starting points and should not 
be identical, but should have displayed equivalent behavior.

Plateaus were only observed for the dataset 
with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although 
the length of each plateau record was slightly longer for {\sc flockture} and on average had more runs converging on
the largest plateau (Supp. Table 1). Multiple runs of {\sc flockture} demonstrated that the differences are well within the
variability that we would expect from randomizing the initial grouping of individuals to 
initiate the algorithms.

@@

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well differentiated populations. Infrequently, at higher levels of population differentiation the 
NA-NC and NA-C models in {\sc structure} could be trapped in a local mode resulting in the 
assignment of individuals from two groups into one while individuals from another 
were separated into two populations (Fig.~\ref{fig:FvSdistruct}  Simulation 1). This phenomena 
(only present in the microsatellite datasets) disappears as population differentiation 
declines. As the subtlety of genetic structure 
increased loss for both programs increases and differences among markers emerged.
For the microsatellite dataset we observed that both programs
produce roughtly equivalent results at higher levels of population differentiation. At intermediate 
values, the variance of the loss incurred by each model increases and certain models do marginally
better in assignment. No model does consistently better than the others, until $F_{ST}$ drops 
below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
 {\sc flockture} did, however, have consistenlty smaller loss than  {\sc structure} at high to intermediate 
values of $F_{ST}$. Once $F_{ST}$ fell below 0.016 {\sc structure} began correctly assigning a 
higher proportion of individuals.
